Staged Abstract Interpreters

### 2018-11-02 paper/20_bg.tex
fix bug / revision

As higher-order functions had been widely introduced in mainstream programming languages \todo{cite}, in this paper, 
we target a tiny higher-order functional language \TLang, which is based on direct-style lambda-calculus, 
and with the support of numbers, arithmetics, recursions and conditionals.
In section \ref{cases_imp}, we will see adding more imperative features to the language.
Since we are mostly interested in analyzing the dynamic behavior of the program, the static semantics and
type system will be omitted. We also assume all the input programs are well-typed.
The abstract syntax is described below:

@@

\begin{lstlisting}
trait Semantics[R[_]] {
  type Ident;    type Addr;   type Value;    type Ans
  type Env <: EnvSpec[R, Ident, Addr, Env]
  type Store <: StoreSpec[R, Addr, Value, Store]
  val $\rho$0: Env;     val $\sigma$0: Store
  def alloc(x: Ident, $\sigma$: Store): R[Addr]
  type EvalFun = (Expr, Env, Store) => Ans
  def eval(ev: EvalFun)(e: Expr, $\rho$: Env, $\sigma$: Store): Ans
  def fix(ev: EvalFun => EvalFun): EvalFun = 
    (e, $\rho$, $\sigma$) =>  ev(fix(ev))(e, $\rho$, $\sigma$)
  def eval_top(e: Expr): Ans = fix(eval)(e, $\rho$0, $\sigma$0)
}
\end{lstlisting}

@@

Readers may notice an additional notation \texttt{[R[\_]]} afterward the trait definition,
which indicates \texttt{Semantics} takes a type parameter \texttt{R}, where \texttt{R} represents
a higher-kinded type and will used to specify the binding-time, i.e., staged or not. 
For the moment, \texttt{R} will be instantiated unstaged --- an identity type constructor: 
\texttt{type NoRep[A] = A}. \todo{where talk about stage poly?}

@@

\begin{lstlisting}
trait Concrete[R[_]] extends Semantics[R] {
  type Ident = String;         type Addr = Int
  type Ans = (Value, Store);   abstract class Value
  case class CloV($\lambda$: Lam, $\rho$: Env) extends Value
  case class NumV(i: Int) extends Value
}
\end{lstlisting}

@@

\begin{lstlisting}
object NoRepConcreteEval extends Concrete[NoRep] {
  def alloc(x: Ident, $\sigma$: Store): Addr = $\sigma$.size + 1
  def eval(ev: EvalFun)(e: Expr, $\rho$: Env, $\sigma$: Store): Ans = 
    e match {
      case Lit(i) => (NumV(i), $\sigma$)
      case Var(x) => ($\sigma$($\rho$(x)), $\sigma$)
      case Lam(x, e) => (CloV(Lam(x,e), $\rho$), $\sigma$)
      case App(e1, e2) =>
        val (CloV(Lam(x,e), $\lambda\rho$), e1$\sigma$) = ev(e1, $\rho$, $\sigma$)
        val (e2v, e2$\sigma$) = ev(e2, $\rho$, e1$\sigma$)
        val $\alpha$ = alloc(x, e2$\sigma$)
        ev(e, $\lambda\rho$ + (x -> $\alpha$), e2$\sigma$ + ($\alpha$ -> e2v))
      case If0(cnd, thn, els) => ...
      case AOp(op, e1, e2) => ...
      case Rec(x, e, body) => ...
    }
}
\end{lstlisting}

@@

\begin{lstlisting}
trait Abstract[R[_]] extends Semantics[R] {
  type Ident = String;         case class Addr(x: Ident)
  type Ans = (Value, Store);   type Value = Set[AbsValue]
  abstract class AbsValue
  case class CloV($\lambda$: Lam, $\rho$: Env) extends AbsValue
  case class NumV() extends AbsValue
}
\end{lstlisting}

>>>>

As higher-order functions had been widely introduced in mainstream programming languages \todo{cite}, in this paper, 
we target a tiny higher-order functional language \TLang, which is based on direct-style lambda-calculus
with the support of numbers, arithmetics, recursions and conditionals.
In section \ref{cases_imp}, we will see adding more imperative features to the language.
Since we are mostly interested in analyzing the dynamic behavior of the program, the static semantics and
type system will be omitted. We also assume all the input programs are well-typed.
The abstract syntax is described below:

@@

\begin{lstlisting}
trait Semantics {
  type R[+T]
  type Ident;    type Addr;   type Value;    type Ans
  type Env <: EnvSpec[R, Ident, Addr, Env]
  type Store <: StoreSpec[R, Addr, Value, Store]
  val $\rho$0: Env;     val $\sigma$0: Store
  def alloc(x: Ident, $\sigma$: Store): R[Addr]
  type EvalFun = (Expr, Env, Store) => Ans
  def eval(ev: EvalFun)(e: Expr, $\rho$: Env, $\sigma$: Store): Ans
  def fix(ev: EvalFun => EvalFun): EvalFun = 
    (e, $\rho$, $\sigma$) =>  ev(fix(ev))(e, $\rho$, $\sigma$)
  def eval_top(e: Expr): Ans = fix(eval)(e, $\rho$0, $\sigma$0)
}
\end{lstlisting}

@@

Readers may notice an additional abstract type \texttt{R[+T]} is defined in the trait,
which means \texttt{R} represents a higher-kinded type and will used to specify the binding-time, 
i.e., staged or not. Since the LMS framework we are using allows us to specify binding-time purely
based on types, the unstaged higher-kinded type can be simply instantiated as an identity type constructor: 
\texttt{type NoRep[+T] = T}. \todo{where talk about stage poly?}

@@

\begin{lstlisting}
trait Concrete extends Semantics {
  type Ident = String;         type Addr = Int
  type Ans = (Value, Store);   abstract class Value
  case class CloV($\lambda$: Lam, $\rho$: Env) extends Value
  case class NumV(i: Int) extends Value
}
\end{lstlisting}

@@

\begin{lstlisting}
object NoRepConcreteEval extends Concrete {
  type R[+T] = NoRep[T]
  def alloc(x: Ident, $\sigma$: Store): Addr = $\sigma$.size + 1
  def eval(ev: EvalFun)(e: Expr, $\rho$: Env, $\sigma$: Store): Ans = 
    e match {
      case Lit(i) => (NumV(i), $\sigma$)
      case Var(x) => ($\sigma$($\rho$(x)), $\sigma$)
      case Lam(x, e) => (CloV(Lam(x,e), $\rho$), $\sigma$)
      case App(e1, e2) =>
        val (CloV(Lam(x,e), $\lambda\rho$), e1$\sigma$) = ev(e1, $\rho$, $\sigma$)
        val (e2v, e2$\sigma$) = ev(e2, $\rho$, e1$\sigma$)
        val $\alpha$ = alloc(x, e2$\sigma$)
        ev(e, $\lambda\rho$ + (x -> $\alpha$), e2$\sigma$ + ($\alpha$ -> e2v))
      case If0(cnd, thn, els) => ...
      case AOp(op, e1, e2) => ...
      case Rec(x, e, body) => ...
    }
}
\end{lstlisting}

@@

\begin{lstlisting}
trait Abstract extends Semantics {
  type Ident = String;         case class Addr(x: Ident)
  type Ans = (Value, Store);   type Value = Set[AbsValue]
  abstract class AbsValue
  case class CloV($\lambda$: Lam, $\rho$: Env) extends AbsValue
  case class NumV() extends AbsValue
}
\end{lstlisting}
